### 2014-09-22 text/Chapter07.tex
Starting the final chapter (Chapter 7)

\chapter{Conclusions} 
\label{ch:conclusions}

%---------------------------------------------------------------------------------------------------

How to make this practical?

codo notation / computation expressions

embedding algebraic things

%---------------------------------------------------------------------------------------------------

@@

%===================================================================================================

@@

\section{Introduction}

\begin{figure}
\begin{equation*}
\tyrule{var}
  {x : \alpha \in \Gamma}
  {\Gamma \vdash x : \alpha }
\end{equation*}
\begin{equation*}
\tyrule{app}
  {\Gamma \vdash e_1 : \alpha \rightarrow \beta &
   \Gamma \vdash e_2 : \alpha }
  {\Gamma \vdash e_1~e_2 : \beta}
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\Gamma, x:\alpha \vdash e : \beta}
  {\Gamma \vdash \lambda x.e : \alpha \rightarrow \beta }
\end{equation*}
\begin{equation*}
\tyrule{letbox}
  {\Gamma \vdash e_1 : \ctyp{r \mseq s}{\alpha} & \Gamma, x : \ctyp{r}{\alpha} \vdash e_2 : \beta }
  {\Gamma \vdash \kvd{let~box}~x=e_1~\kvd{in}~e_2 : \ctyp{s}{\beta}}
\end{equation*}
\begin{equation*}
\tyrule{eval}
  {\Gamma \vdash e : \ctyp{\munit}{\alpha}}
  {\Gamma \vdash !e : \alpha}
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\Gamma \vdash e : \ctyp{s}{\alpha} }
  {\Gamma \vdash e : \ctyp{r}{\alpha} }
\;(s\leq r)  
\end{equation*}

\caption{Type system for the coeffect meta-language cml}
\label{fig:cml-types}
\end{figure}

@@

\section{Summary}

>>>>

\chapter{Further work and conclusions} 
\label{ch:conclusions}

The main goal of this thesis is to demonstrate that there is a unified notion of \emph{context}
that captures many common scenarios in computer programming. The examples we explore include
many notions of context important in modern software development (including distributed and
cross-platform development) as well as context studied in fundamental programming language 
research (including data-flow, linear logics and variable liveness).

This thesis demonstrates that all these notions of context have something in common. In 
particular, they can be modelled by our flat or structural coeffect calculi. Although the 
results presented in this thesis are mainly of a theoretical nature, we, indeed, believe that 
coeffects should be integrated in main-stream programming languages.

In this chapter, we present further work in two directions. Firstly, we outline one possible
approach for practical implementations of coeffects (Section~\ref{sec:conclusions-impl}). 
Secondly, we discuss an alternative approach to defining coeffect systems which highlights
the relationship between our work and related work arising from modal logics 
(Section~\ref{sec:conclusions-meta}). Finally, the Section~\ref{sec:conclusions-summary} summarizes
and concludes the thesis.

% =================================================================================================
%                                                                   
%      ####                         #       #                   ##   
%      #   #                        #                            #   
%      #   #  # ##    ###    ###   ####    ##     ###    ###     #   
%      ####   ##  #      #  #   #   #       #    #   #      #    #   
%      #      #       ####  #       #       #    #       ####    #   
%      #      #      #   #  #   #   #  #    #    #   #  #   #    #   
%      #      #       ####   ###     ##    ###    ###    ####   ###  
%                                                                
% =================================================================================================

\section{Towards practical coeffects}
\label{sec:conclusions-impl}

As discussed earlier, the main focus of this thesis is the development of the much needed 
\emph{theory of context-aware computations} and so discussing the details of a practical 
implementation of coeffect tracking is beyond the scope of the thesis. However, this section
briefly outlines one possible pathway towards this goal.

Many of the examples of contextual computation that we discussed earlier (\eg~implicit
parameters \cite{app-implicit-parameters} or distributed computations 
\cite{app-distributed-ml5,app-distributed-links}) have been implemented as a single-purpose
programming language feature. However, the main contribution of this thesis is that it captures
\emph{multiple} different notions of context-aware computations using just a \emph{single}
common abstraction. For this reason, we advocate that future practical implementations of
coeffects should not be single-purpose language features, but instead reusable abstractions that
can be instantiated with a concrete \emph{coeffect algebra} specified by the user.

In order to do this, programming languages need to provide two features; one that allows 
embedding of context-aware computations themselves in programs (akin to the ``do'' notation
in Haskell) and one that allows tracking of the contextual information in the type system.

% -------------------------------------------------------------------------------------------------

\subsection{Embedding contextual computations}

The embedding of \emph{contextual} computations in programming languages can learn from better 
explored emedding of \emph{effectful} computations. In purely functional programming languages
such as Haskell, effectful computations are embedded by \emph{implementing} them and inserting the
necessary (monadic) plumbing. This is made easier by the ``do'' notation that inserts the monadic
operations automatically.

The recently proposed ``codo'' notation \cite{comonads-codo} provides a similar plumbing for
context-aware computations based on comonads. The notation is close to the semantics of our 
flat coeffect calculus (Chapter~\ref{ch:flat}). Extending it to support calculi based on the
structural coeffect calculus (Chapter~\ref{ch:structural}) is an interesting futrue work -- 
this requires explicitly manipulating individual context variables and application of structural
rules (which is not needed in flat coeffects).

In ML-like languages, effects (and many coeffects) are built-in into the language semantics, 
but they can still use a special notation for explicitly marking effectful (coeffectful) 
blocks of code. In F\#, this is done using \emph{computation expressions}  \cite{app-computation-zoo}
that differ from the ``do'' notation in two ways. First, they support wider range of 
constructs, making it possible to wrap existing F\# code in a block without other changes.
Second, they support other abstractions including monads, applicative functors and monad
transformers. It would be interesting to see if computation expressions can be extended to
handle computations based on flat/structural indexed comonads.

Finally, more lightweight syntax for effectful computation can be obtained using
techniques that automatically insert the necessary monadic plumbing (wihtout using any
special syntax). This has been done in the context of effecful computations \cite{monads-lightweight-ml}
and similar approach would be worth exploring for coeffects.

% -------------------------------------------------------------------------------------------------

\subsection{Coeffect annotations as types}

The other aspect of practical implementation of coeffects is tracking the context requirements
(coeffect annotations) in the type system. To achieve this, the type system needs to be able to 
capture the different examples of coeffect algebras. The structures used in this thesis include sets 
(with union or interesection), natural numbers (with addition, maximum, minimum and multiplication),
two-point lattice (for liveness) and free monoids (vectors of annotations).

A recent work on embedding effect systems in Haskell \cite{effects-embedding} shows that the
recent additions to the Haskell type system are powerful enough to implement structures such
as sets at the type level.

\cite{types-dependent-ml}
\cite{types-units-of-measure}
\cite{app-inforich}

% =================================================================================================
%                                                                                             
%    #   #          #                    ##                                                    
%    #   #          #                     #                                                    
%    ## ##   ###   ####    ###            #     ###   # ##    ## #  #   #   ###    ## #   ###  
%    # # #  #   #   #         #  #####    #        #  ##  #  #  #   #   #      #  #  #   #   # 
%    #   #  #####   #      ####           #     ####  #   #   ##    #   #   ####   ##    ##### 
%    #   #  #       #  #  #   #           #    #   #  #   #  #      #  ##  #   #  #      #     
%    #   #   ###     ##    ####          ###    ####  #   #   ###    ## #   ####   ###    ###  
%                                                            #   #                #   #        
%                                                             ###                  ###         
% =================================================================================================

Related and further work has been discussed throughout, but one important thing remains.

@@

@@

%\begin{figure}
\begin{equation*}
\tyrule{var}
  {x : \alpha \in \Gamma}
  {\Gamma \vdash x : \alpha }
\end{equation*}
\begin{equation*}
\tyrule{app}
  {\Gamma \vdash e_1 : \alpha \rightarrow \beta &
   \Gamma \vdash e_2 : \alpha }
  {\Gamma \vdash e_1~e_2 : \beta}
\end{equation*}
\begin{equation*}
\tyrule{abs}
  {\Gamma, x:\alpha \vdash e : \beta}
  {\Gamma \vdash \lambda x.e : \alpha \rightarrow \beta }
\end{equation*}
\begin{equation*}
\tyrule{letbox}
  {\Gamma \vdash e_1 : \ctyp{r \mseq s}{\alpha} & \Gamma, x : \ctyp{r}{\alpha} \vdash e_2 : \beta }
  {\Gamma \vdash \kvd{let~box}~x=e_1~\kvd{in}~e_2 : \ctyp{s}{\beta}}
\end{equation*}
\begin{equation*}
\tyrule{eval}
  {\Gamma \vdash e : \ctyp{\munit}{\alpha}}
  {\Gamma \vdash !e : \alpha}
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\Gamma \vdash e : \ctyp{s}{\alpha} }
  {\Gamma \vdash e : \ctyp{r}{\alpha} }
\;(s\leq r)  
\end{equation*}

%\caption{Type system for the coeffect meta-language cml}
%\label{fig:cml-types}
%\end{figure}

@@

\section{Summary}
\label{sec:conclusions-summary}
